---
title: "Gridding point data by Kriging"
author: "Marko Laine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  \VignetteIndexEntry{Gridding point data by Kriging}
  \VignetteEngine{knitr::rmarkdown}
  \VignetteEncoding{UTF-8}
---

This document describes grdding of point observations to a regular grid. The code is intended to grid station temperatures (or other meterological variables) to a grid. The code can use a prior backgroud model.

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Examples

```{r, eval=FALSE, include=TRUE}
library(MOSfieldutils)
MOSgrid(stationsfile = stationdata,modelgridfile = mosgrid, bgfile = ECbg)
```


```{r, eval=FALSE, include=TRUE}
plot_MOS_field(mosfield)
```


## Code details

The essential parts of the fortran code is the following

```fortran
  call chol(b, nobs)               ! replace B with L = chol(B,’lower')
  call ldivide(b,x, nobs, npar)    ! replace x with L\x
  call ldivide(b,y, nobs, 1)       ! replace y with L\y
  call lsq(x,y,beta, nobs, npar)   ! beta = x\y
  call resid(x,y,beta, nobs, npar) ! y = y - x*beta
  call ltdivide(b,y, nobs, 1)      ! y = L'\y
  ! calculate prediction over grid
  do i=1, ngrid
     x1 = grid(i,:)
     call corrdist(cgrid(i,:),cy,b1, nobs, covpars)
     ypred(i) = matmul(x1,beta) + matmul(b1,y)
  end do
```

## Mathematical background 

Basically, the gridding can be described

\begin{equation}
  \label{equ:hiermodel}
  \begin{split}
  y|\eta &\sim N(H\eta,\Sigma_y), \qquad\text{observations}\\
  \eta|\beta &\sim N(X\beta,\Sigma_\eta), \qquad\text{spatial random field}\\
  \beta &\sim N(\beta_0,\Sigma_\beta)  \qquad\text{hyper parameters}
  \end{split}
\end{equation}


## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette. 

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
