---
title: "Gridding point data by Kriging"
author: "Marko Laine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette: fig_caption: yes
vignette: >
  \VignetteIndexEntry{Gridding point data by Kriging}
  \VignetteEngine{knitr::rmarkdown}
  \VignetteEncoding{UTF-8}
---

DUMMY text, still, sorry.

This document describes grdding of point observations to a regular grid. The code is intended to grid station temperatures (or other meterological variables) to a grid. The code can use a prior backgroud model. The code was done as the default Kriging in the pakage gstat was so slow.


## Examples

```{r, eval=FALSE, include=TRUE}
library(MOSfieldutils)
MOSgrid(stationsfile = stationdata,modelgridfile = mosgrid, bgfile = ECbg)
```


```{r, eval=FALSE, include=TRUE}
plot_MOS_field(mosfield)
```


## Code details

The essential parts of the fortran code is the following

```fortran
  call chol(b, nobs)               ! replace B with L = chol(B,’lower')
  call ldivide(b,x, nobs, npar)    ! replace x with L\x
  call ldivide(b,y, nobs, 1)       ! replace y with L\y
  call lsq(x,y,beta, nobs, npar)   ! beta = x\y
  call resid(x,y,beta, nobs, npar) ! y = y - x*beta
  call ltdivide(b,y, nobs, 1)      ! y = L'\y
  ! calculate prediction over grid
  do i=1, ngrid
     x1 = grid(i,:)
     call corrdist(cgrid(i,:),cy,b1, nobs, covpars)
     ypred(i) = matmul(x1,beta) + matmul(b1,y)
  end do
```

## Mathematical background 

Basically, the gridding can be described as

\begin{equation}
  \label{equ:hiermodel}
  \begin{split}
  y|\eta &\sim N(H\eta,\Sigma_y), \qquad\text{observations}\\
  \eta|\beta &\sim N(X\beta,\Sigma_\eta), \qquad\text{spatial random field}\\
  \beta &\sim N(\beta_0,\Sigma_\beta)  \qquad\text{hyper parameters}
  \end{split}
\end{equation}


## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold',fig.cap="Capttion"}
MOS_plot_field(out)
```


